Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-06 rnni_convexity.tex
intro for Geometry of RNNI section

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}
\newcommand{\mdtree}{\textsc{MDTree}}

@@

In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
They enable us to prove that between caterpillar trees $T$ and $R$ there is always a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version, the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}).

Before we get into details of these geometric properties of $\rnni$, we need to prove the following auxiliary results.

\begin{lemma}
It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
\label{lemma:diameter_bound}
\end{lemma}

@@

\begin{corollary}
The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{corollary:radius}
\end{corollary}

>>>>

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}
\newcommand{\mdtree}{\textsc{MDTree}}
\newcommand{\rad}{\textsc{R}}

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
We will give a counterexample to the Split Theorem in $\rnni$ and consider the Weak Split Theorem that states that every shared split of taxa if maintained along a shortest path (see Conjecture~\ref{conjecture:cluster_theorem}).

We start with the following auxiliary results.

\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
\label{lemma:diameter_bound}
\end{lemma}

@@

\begin{corollary}
The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{corollary:radius}
\end{corollary}
